---
title: "DC for Social Science: Exercises for Day 2"
author: "Lachlan Deer"
output: 
  unilur::tutorial_pdf_solution: default
  unilur::tutorial_pdf: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = TRUE, eval=TRUE, message=FALSE}
library(dplyr)
library(lubridate)
library(tidyr)
library(ggplot2)
```


```{r, echo = FALSE, eval=TRUE, message=FALSE, purl=FALSE}
dest_file <- "data/SAFI_clean.csv"
if(!file.exists(dest_file)){
  dir.create('data')
  download.file("https://ndownloader.figshare.com/files/11492171", destfile =  dest_file, mode = "wb")
} else {
message ('data already downloaded')  
}
```

```{r, echo = TRUE, eval=TRUE, message=FALSE}
interviews <- readr::read_csv('data/SAFI_clean.csv', na = "NULL")

```

# Working With Data

## Exercise 1

- Create a data frame (`interviews_100`) containing only the data in row 100 of the `interviews` dataset.

```{r, solution = TRUE}
interviews_100 <- interviews[100, ]
```

- Notice how `nrow()` gave you the number of rows in a data frame?
  - Use that number to pull out just that last row in the data frame.
  - Compare that with what you see as the last row using `tail()` to make sure it's meeting expectations.
  - Pull out that last row using `nrow()` instead of the row number.
  - Create a new data frame (`interviews_last`) from that last row.

```{r, solution = TRUE}
n_rows <- nrow(interviews)
interviews_last <- interviews[n_rows, ]
```

- Using the number of rows in the interviews dataset that you found in  question 2, extract the row that is in the middle of the dataset. Store the content of this middle row in an object named `interviews_middle`. (hint: This dataset has an odd number of rows, so finding the middle is a bit trickier than dividing n_rows by 2. Use the median( ) function and what you've learned about sequences in R to extract the middle row!)

```{r, solution = TRUE}
interviews_middle <- interviews[median(1:n_rows), ]
```

- Combine nrow() with the - notation above to reproduce the behavior of head(interviews), keeping just the first through 6th rows of the interviews dataset

```{r, solution = TRUE}
interviews_head <- interviews[-(7:n_rows), ]
```

## Exercise 2

- Using pipes, subset the interviews data to include interviews where respondents were members of an irrigation association (`memb_assoc`) and retain only the columns `affect_conflicts`, `liv_count`, and `no_meals`.

```{r, solution = TRUE}
interviews %>%
    filter(memb_assoc == "yes") %>%
    select(affect_conflicts, liv_count, no_meals)
```

- Create a new dataframe from the interviews data that meets the following criteria: contains only the village column and a new column called total_meals containing a value that is equal to the total number of meals served in the household per day on average (no_membrs times no_meals). Only the rows where total_meals is greater than 20 should be shown in the final dataframe.

```{r, solution = TRUE}
interviews_total_meals <- interviews %>%
    mutate(total_meals = no_membrs * no_meals) %>%
    filter(total_meals > 20) %>%
    select(village, total_meals)
```

- Use group_by() and summarize() to find the mean, min, and max number of household members for each village. Also add the number of observations (hint: see ?n).

```{r, solution = TRUE}
interviews %>%
  group_by(village) %>%
  summarize(
      mean_no_membrs = mean(no_membrs),
      min_no_membrs = min(no_membrs),
      max_no_membrs = max(no_membrs),
      n = n()
  )
```

- What was the largest household interviewed in each month?

```{r, solution = TRUE}
interviews %>%
    mutate(month = month(interview_date),
           year = year(interview_date)) %>%
    group_by(year, month) %>%
    summarize(max_no_membrs = max(no_membrs))
```

## Exercise 3

- Create a new dataframe (named interviews_months_lack_food) that has one column for each month and records `TRUE` or `FALSE` for whether each interview respondent was lacking food in that month.

```{r, solution = TRUE}
interviews_months_lack_food <- interviews %>%
  separate_rows(months_lack_food, sep = ";") %>%
  mutate(months_lack_food_logical  = TRUE) %>%
  pivot_wider(names_from = months_lack_food, 
              values_from = months_lack_food_logical, 
              values_fill = list(months_lack_food_logical = FALSE))
```


- How many months (on average) were respondents without food if they did belong to an irrigation association? What about if they didn't?

```{r, solution = TRUE}
interviews_months_lack_food %>%
  mutate(number_months = rowSums(select(., Jan:May))) %>%
  group_by(memb_assoc) %>%
  summarize(mean_months = mean(number_months))
```

# Plotting

```{r, echo = FALSE}
interviews_plotting <- interviews %>%
  ## pivot wider by items_owned
  separate_rows(items_owned, sep = ";") %>%
  ## if there were no items listed, changing NA to no_listed_items
  replace_na(list(items_owned = "no_listed_items")) %>%
  mutate(items_owned_logical = TRUE) %>%
  pivot_wider(names_from = items_owned, 
              values_from = items_owned_logical, 
              values_fill = list(items_owned_logical = FALSE)) %>%
  ## pivot wider by months_lack_food
  separate_rows(months_lack_food, sep = ";") %>%
  mutate(months_lack_food_logical = TRUE) %>%
  pivot_wider(names_from = months_lack_food, 
              values_from = months_lack_food_logical, 
              values_fill = list(months_lack_food_logical = FALSE)) %>%
  ## add some summary columns
  mutate(number_months_lack_food = rowSums(select(., Jan:May))) %>%
  mutate(number_items = rowSums(select(., bicycle:car)))
```

## Exercise 1

Boxplots are useful summaries, but hide the shape of the distribution. For example, if the distribution is bimodal, we would not see it in a boxplot. An alternative to the boxplot is the violin plot, where the shape (of the density of points) is drawn.

- Replace the box plot with a violin plot; see geom_violin().

```{r, solution = TRUE}
interviews_plotting %>%
  ggplot(aes(x = respondent_wall_type, y = rooms)) +
  geom_violin(alpha = 0) +
  geom_jitter(alpha = 0.5, color = "tomato")
```

- Create a boxplot for `liv_count` for each wall type. Overlay the boxplot layer on a jitter layer to show actual measurements.

```{r, solution = TRUE}
interviews_plotting %>%
   ggplot(aes(x = respondent_wall_type, y = liv_count)) +
   geom_boxplot(alpha = 0) +
   geom_jitter(alpha = 0.5, width = 0.2, height = 0.2)
```
 
- Add colour to the data points on your boxplot according to whether the respondent is a member of an irrigation association (memb_assoc).

```{r, solution = TRUE}
interviews_plotting %>%
  ggplot(aes(x = respondent_wall_type, y = liv_count)) +
  geom_boxplot(alpha = 0) +
  geom_jitter(aes(color = memb_assoc), alpha = 0.5, width = 0.2, height = 0.2)
```

## Exercise 2

```{r, echo = TRUE, eval = TRUE}
### Intermezzio - Joining Data
village_pop <- interviews_plotting %>%
        group_by(village) %>%
        count() %>%
        rename(people_in_village = n)

# Example of how to merge data 
interviews_plotting_2 <- interviews_plotting %>%
                  inner_join(village_pop, by = c("village"))
```

- Create a plot that visualizes the proportion of respondents in each village who owned a particular item?


```{r, solution = TRUE}
percent_items <- interviews_plotting %>%
    pivot_longer(cols = bicycle:no_listed_items, names_to = "items", 
                 values_to = "items_owned_logical") %>%
    filter(items_owned_logical) %>%
    count(items, village) %>%
    ## add a column with the number of people in each village
    #mutate(people_in_village = case_when(village == "Chirodzo" ~ 39,
    #                                     village == "God" ~ 43,
    #                                     village == "Ruaca" ~ 49)) %>%
    inner_join(village_pop, by = c("village")) %>%
    mutate(percent = (n / people_in_village) * 100)

percent_items %>%
    ggplot(aes(x = village, y = percent)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~ items) +
    theme_bw() +
    theme(panel.grid = element_blank())
```


